Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2018-01-25 writeup/response2reviews.tex

\textbf{Reviewer 1}\\
The paper introduces a model-based method called conStruct to account for isolation-by-distance (IBD) when clustering individuals into populations. IBD is known to generate spurious clusters and the proposed software package can avoid spurious cluster detection. Accounting for space in the proposed model based framework amounts at introducing a regularization penalty when fitting admixture model. Regularization explains why overfitting is avoided and correct admixture proportions are found even when performing model-based clustering with a value of K that is too large.

This being said, I have serious concerns about the submitted manuscript.

\begin{itemize}
\item First, many results concern comparisons between clustering based on the spatial and non-spatial models of construct (Figs 2, 3, 4, 6, 7, 9). However the authors should rather compare their spatial approach to existing software for model-based clustering, which do not account for space (Structure, fastSTRUCTURE, ADMIXTURE,...). Results obtained with one of these software should not be available in supplementary material but should rather be visible in the main text.\\\\
\gb{The utility of the nonspatial model implemented in \texttt{conStruct} 
is that its performance can be directly compared to that of the spatial model; 
their cross-validation predictive accuracy values are in the same currency 
because they use the same Wishart likelihood model.
So, using it we can say not just that that the nonspatial model prefers 
$K=X$ while the spatial model prefers $K=Y$, 
but also, e.g., that the spatial model is preferred over the nonspatial model.
However, as you note, it's important to compare \texttt{conStruct}'s performance to 
that of existing nonspatial model-based clustering methods as well, 
which is why we had also included analyses using fastSTRUCTURE.

\ Your point here is well-taken though, because the fastSTRUCTURE analyses, 
which included no model comparisons, did not provide a direct basis of comparison to \texttt{conStruct}.
There is a model comparison tool implemented in fastSTRUCTURE, 
but we were unable to get it to run on our datasets.
We have contacted the fastSTRUCTURE authors 
and alerted them to this issue, 
but have yet to hear back from them.

\ We have therefore analyzed all datasets included in the paper using ADMIXTURE, 
and used ADMIXTURE's cross-validation procedure to compare model performance across values of $K$.  
These results are discussed extensively throughout the updated manuscript ``Results" section; 
and the model that minimizes the cross-validation error in the ADMIXTURE analyses on simulated data 
is highlighted in the the simulation cross-validation results figure.
ONLY KEEP THIS IN IF WE LIKE THE NEW FIG.
However, due to the already extensive length of the manuscript (which Reviewer 2 points out), 
the results of the ADMIXTURE analyses are given in the Supplementary Materials.
}\\

@@

\item 100: Is there a reason that allelic covariance is used instead of genetic covariance?\\\\
\gb{
    Good question; we've added some more text for this, 
    both explaining what the allelic covariance is and why we use it in lieu of the genetic covariance,
    (Lines XXX-XXX).
    The short answer is that the allelic covariance is insensitive to the choice of reference allele, 
    and, because it doesn't subtract off means like the genetic covariance, 
    has a constant value between a pair of samples in the dataset, 
    regardless of which other samples are included.
}\\

\item Could supp. figures be sorted according to first mentioning in ms?\\\\
\gb{MAKE SURE THIS IS RIGHT AFTER MAKING SUBMISSION MS}\\

\item Fig3: increase label size\\\\
\gb{Fixed.}\\

\item Last line in paragraph on p. 18: as the derivations are otherwise very detailed and easy to follow, it might be worth adding here that this follows from E[UV] = 4 Cov[A, B]\\\\
\gb{
We have substantially rewritten this section for greater clarity.\\
}

@@

\item 634: Using ms,
\\\\
\gb{Fixed.}\\

\item Fig S4: Could the colors be made to correspond each other (e.g. panel e vs 3). In general, is the exchangeability of layers a practical concern, or similar to the original structure, does the MCMC never ``switch'' two layer labels?\\\\
\gb{In STAN, we declare the vector of $\phi$ parameters to be an object of class ``positive ordered",
	meaning that values of phi must all have values greater than zero, 
	and must be in strictly increasing order.
	In practice, I don't think that much label-switching would occur within the course of a single MCMC run 
	(like in STRUCTURE), but inducing a specific order to the $\phi$ parameter makes that less likely still.
	There can be label switching across independent runs for the same value of $K$, 
	and of course there's label switching across runs with different numbers of $K$.
	In the \texttt{conStruct} R package, we've included a function that uses the cluster matching 
	algorithm described in CLUMPP (Jakobsson and Rosenberg 2007) to match layers across runs.
	We used that algorithm to keep plotting colors consistent in our figures, including in Fig. XXXX.
	However, something funny happens in that sequence of analyses between $K=5$ and $K=6$, 
	where the ranking of cluster matching algorithm gives wonky results, 
	leading to the non-intuitive layer color matching.
	We've tweaked that one by hand to keep the plotting nice.
 }\\

### 2018-01-28 writeup/response2reviews.tex

\item Third, I think that the main point of the paper should be more explicit. Non-spatial model based approaches are prone to overfitting when K is too large and this can be avoided when using spatial-based approaches. \\\\
\gb{
Good suggestion; we've added a line to the Abstract to this effect.
}\\

\item Last, there are other software that account for space when performing model-based clustering and they should be included in the comparisons (BAPS, Geneland, TESS,...). Right now, they are only mentioned in the introduction and we would like to see results comparing these different model-based approaches.\\\\
\gb{
The methods mentioned here --- BAPS, Geneland, and TESS --- 
place spatial priors on admixture proportions
and/or cluster membership, 
but like STRUCTURE, they assume that allele frequencies 
within a cluster are constant across space.
Therefore, although spatially explicit, this parameterization does not 
account for space in a way that avoids the problem we describe in our manuscript: 
the characterization of spatially continuous differentiation 
as clines of admixture across spurious clusters.

\ A comprehensive comparison between \texttt{conStruct} 
and all other model-based clustering methods is beyond the scope of this paper. 
As STRUCTURE is the canonical model-based clustering algorithm
(having been cited $>$ 20,000 times), but is too slow to be run 
on the modern-scale SNP datasets we simulate, 
we feel that ADMIXTURE, which implements an extremely similar model 
within a faster inference framework, 
provides the best comparison to our method.}\\

\item In the introduction, the authors write ``placing spatial priors on cluster memberships'' when referring to other methods that deal with ancestry proportions instead of cluster memberships.\\\\
\gb{
Guillot \emph{et. al}, in the 2005 paper that introduced the method Geneland, 
refer to this quantity as ``population membership."
We kept the wording consistent for clarity, but changed ``population" 
to ``cluster" to keep a distinction between biological and statistical groupings.}\\

@@

\item Another thing that could be added are some additional simulations in more complex demographic models, such as expansions, temporally/spatially heterogeneous migration, or sparser sampling schemes. However, as the paper is already rather long (at almost 20 single-spaced pages), and the space of scenarios for which the model is relevant is enormous, I hope the authors consider further exploring their model in a follow-up study.\\\\
\gb{We completely agree, and further explorations of the behavior of the method 
in different scenarios are being planned now.}\\

\item 100: Is there a reason that allelic covariance is used instead of genetic covariance?\\\\
\gb{
    Good question; we've added some more text for this, 
    both explaining what the allelic covariance is and why we use it in lieu of the genetic covariance,
    (in the ``Data" subsection of the ``Results" 
    and the ``Allelic covariance" subsection of ``Materials and Methods").
    The short answer is that the allelic covariance is insensitive to the choice of reference allele, 
    and, because it doesn't subtract off means like the genetic covariance, 
    has a constant value between a pair of samples in the dataset, 
    regardless of which other samples are included.
}\\

@@

\item 634: Using ms,
\\\\
\gb{Fixed.}\\

\item Fig S4: Could the colors be made to correspond each other (e.g. panel e vs 3). In general, is the exchangeability of layers a practical concern, or similar to the original structure, does the MCMC never ``switch'' two layer labels?\\\\
\gb{In STAN, we declare the vector of $\phi$ parameters to be an object of class ``positive ordered",
	meaning that values of phi must all have values greater than zero, 
	and must be in strictly increasing order.
	In practice, I don't think that much label-switching would occur within the course of a single MCMC run 
	(like in STRUCTURE), but inducing a specific order to the $\phi$ parameter makes that less likely still.
	There can be label switching across independent runs for the same value of $K$, 
	and of course there's label switching across runs with different numbers of $K$.
	In the \texttt{conStruct} R package, we've included a function that uses the cluster matching 
	algorithm described in CLUMPP (Jakobsson and Rosenberg 2007) to match layers across runs.
	We used that algorithm to keep plotting colors consistent in our figures, including in Fig. XXXX.
	However, something funny happens in that sequence of analyses between $K=5$ and $K=6$, 
	where the ranking of cluster matching algorithm gives wonky results, 
	leading to the non-intuitive layer color matching.
	We've tweaked that one by hand to keep the plotting nice.
 }\\

\item Fig S26: maybe keep Y-axis scale consistent between panels?\\\\
\gb{Fixed.}\\

\item Cross-Validation: As we all know SNP are not uncorrelated. I am wondering if that could influence the CV, as spurious patterns in the testing data might be partially explained by nearby SNP in the training data. Is that a concern?\\\\
\gb{
	This is a good point and it definitely could be a concern.
	We've included a discussion of this in the cross-validation procedure description, 
	as well as a description of the CV procedure that should be used if you know LD between loci.
	But, we also imagine that, for a while at least, the vast majority of \texttt{conStruct} users 
	won't have genomic position information for their SNP data.
}\\

### 2018-01-28 writeup/conStruct.tex
prep for resubmission

A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
the use of clustering or assignment methods may incorrectly ascribe
differentiation due to continuous processes (e.g., geographic isolation by distance)
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This reflects a shortcoming of current methods for inferring and 
visualizing population structure when applied to genetic data
deriving from geographically distributed populations.
Here, we present a statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of two-dimensional population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This thereby explicitly addresses the ``clines versus clusters" problem in 
modeling population genetic variation,
and remedies some of the overfitting that nonspatial models are prone to.
The method produces useful descriptions of structure in genetic relatedness
in situations where separated, geographically distributed populations interact,
as after a range expansion or secondary contact.
We demonstrate the utility of this approach using simulations 
and by applying it to empirical datasets of poplars and black bears in North America.

@@

The spatial model's better fit is reflected by increased predictive accuracy:
as shown in Fig \ref{sim_xvals},
across all models and choices of $K$, the spatial model is correctly preferred over the nonspatial model.
As desired, predictive accuracy of the spatial model increases until the true value of $K$,
and then plateaus or declines
(Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}),
while predictive accuracy of the nonspatial model increases as subsequent clusters are added.
The same holds true for the ADMIXTURE cross-validation results,
in which models that have the largest number of clusters are preferred over all other models,
as shown in Fig \ref{sim_xvals} (vermilion points),
and, in more detail, in Fig \ref{sim_admix_CVerrors}.
% In contrast, the predictive accuracy of the spatial \texttt{conStruct} model
% either plateaus at or declines from the true value of $K$ used to simulate the data 
% (Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}).

@@

\begin{figure}
	\centering
	\ifincludefigs
		{\includegraphics[width=\textwidth]{figs/sims/sim_xvals.pdf}}
	\fi
	\caption{
	Cross-validation results for data simulated under $K=1$, $K=2$, and $K=3$, 
	comparing the spatial and nonspatial \texttt{conStruct} models 
	(in blue and green, respectively) run with $K=1$ through 7.  
	The inset plots zoom in on cross-validation results outlined in the dotted boxes.
    The spatial model shows better model fit at every value of $K$.
    The ADMIXTURE model with the lowest cross-validation error 
    (i.e., the preferred model) is highlighted with an vermilion diamond.
    In all simulations, the preferred ADMIXTURE model 
    was that with the largest number of clusters.
    }\label{sim_xvals}
\end{figure}

@@

Thanks to averaging over choice of alleles,
the within-population allelic variance in sample $i$, $\widehat{Omega_{i,i}}$,
is the variance of a series of Bernoulli($1/2$) draws across loci,
and therefore $\widehat{\Omega}_{i,i} = 1/4$ for every sample $i$.
Averaging over choice of reference allele therefore removes some information 
about factors acting within populations that might otherwise leave signatures in, 
e.g., the genetic covariance, 
such as population size, extent of inbreeding, and history of bottlenecks.  
However, as our model is focused on modeling 
\emph{co}variances between samples as the outcome of some spatial process, 
we count this a minor loss.

@@

\gb{If the spatial coordinates of the loci in the genome are known, 
the training/testing partitioning should be designed to accommodate linkage disequilibrium (LD).
Loci in strong LD do not represent independent instantiations of the coalescent process, 
so if loci from a single linkage block are randomly split into both training and testing partitions, 
the independence of the test in the testing partition will be compromised 
(e.g., the parameters estimated from the training partition 
might be describing process heterogeneity or noise 
in a region of the genome that also has loci included in the testing partition).
If genomic position or LD between SNPs is known, 
the best practice for cross-validation is to make sure that 
no loci in the testing dataset are in strong LD with, 
or spatially proximate to, loci in the testing dataset.}

>>>>

A classic problem in population genetics is the characterization 
of discrete population structure in the presence of 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
the use of clustering or assignment methods may incorrectly ascribe
differentiation due to continuous processes (e.g., geographic isolation by distance)
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This reflects a shortcoming of current methods for inferring and 
visualizing population structure when applied to genetic data
deriving from geographically distributed populations.
Here, we present a statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of two-dimensional population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This thereby explicitly addresses the ``clines versus clusters" problem in 
modeling population genetic variation,
and remedies some of the overfitting to which nonspatial are prone.
The method produces useful descriptions of structure in genetic relatedness
in situations where separated, geographically distributed populations interact,
as after a range expansion or secondary contact.
We demonstrate the utility of this approach using simulations 
and by applying it to empirical datasets of poplars and black bears in North America.

@@

The spatial model's better fit is reflected by increased predictive accuracy:
as shown in Fig \ref{sim_xvals},
across all models and choices of $K$, the spatial model is correctly preferred over the nonspatial model.
As desired, predictive accuracy of the spatial model increases until the true value of $K$,
and then plateaus or declines
(Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}),
while predictive accuracy of the nonspatial model increases as subsequent clusters are added.
The same holds true for the ADMIXTURE cross-validation results,
in which models that have the largest number of clusters are preferred over all other models,
as shown in Fig \ref{sim_xvals} (vermilion diamonds),
and, in more detail, in Fig \ref{sim_admix_CVerrors}.
% In contrast, the predictive accuracy of the spatial \texttt{conStruct} model
% either plateaus at or declines from the true value of $K$ used to simulate the data 
% (Figs \ref{sim_xvals}, \ref{simK1_xval}, \ref{simK2_xval}, and \ref{simK3_xval}).

@@

\begin{figure}
	\centering
	\ifincludefigs
		{\includegraphics[width=\textwidth]{figs/sims/sim_xvals.pdf}}
	\fi
	\caption{
	Cross-validation results for data simulated under $K=1$, $K=2$, and $K=3$, 
	comparing the spatial and nonspatial \texttt{conStruct} models 
	(in blue and green, respectively) run with $K=1$ through 7.  
	The inset plots zoom in on cross-validation results outlined in the dotted boxes.
    The spatial model shows better model fit at every value of $K$.
    The vermillion diamond indicates the value of $K$ selected 
    on the basis of lowest cross-validation error among ADMIXTURE models. 
    In all simulations, the preferred ADMIXTURE model 
    was that with the largest number of clusters.
    }\label{sim_xvals}
\end{figure}

@@

Thanks to averaging over choice of alleles,
the within-population allelic variance in sample $i$, $\widehat{\Omega}_{i,i}$,
is the variance of a series of Bernoulli($1/2$) draws across loci,
and therefore $\widehat{\Omega}_{i,i} = 1/4$ for every sample $i$.
Averaging over choice of reference allele therefore removes some information 
about factors acting within populations that might otherwise leave signatures in, 
e.g., the genetic covariance, 
such as population size, extent of inbreeding, and history of bottlenecks.  
However, as our model is focused on modeling 
\emph{co}variances between samples as the outcome of some spatial process, 
we count this a minor loss.

@@

If the spatial coordinates of the loci in the genome are known, 
the training/testing partitioning should be designed to accommodate linkage disequilibrium (LD).
Loci in strong LD do not represent independent instantiations of the coalescent process, 
so if loci from a single linkage block are randomly split into both training and testing partitions, 
the independence of the test in the testing partition will be compromised 
(e.g., the parameters estimated from the training partition 
might be describing process heterogeneity or noise 
in a region of the genome that also has loci included in the testing partition).
If genomic position or LD between SNPs is known, 
the best practice for cross-validation is to make sure that 
no loci in the testing dataset are in strong LD with, 
or spatially proximate to, loci in the testing dataset.
